### 2014-09-06 relax1.tex
intro+relaxI

\section{Relaxation of Flow Dependencies}~\label{sec:relax1}
Given the trace, $\tau=<\Gamma , \{\tau_{t_1}, \tau_{t_2}, \dots \tau_{t_n} \}, O, \theta>$, the main goal of our analysis is to derive a new trace,  $\tau'=<\Gamma , \{\tau_{t_1}, \tau_{t_2}, \dots \tau_{t_n} \}, O', \theta'>$, which reschedules the events from different threads (the order of events inside each thread should remain unchanged) to witness the race. The key insight underlying our first relaxation is that, the read events in $\tau'$ are allowed by our analysis to read different values from different writes (than in trace $\tau$), but they are required by existing approaches to read the same values as in $\tau$. Therefore, our relaxation allows more schedules, which are likely to expose more races. Note that we also need to compute the new value mapping $\theta'$ because the rescheduling may alter the value every variable is assigned to. The computation of the new schedule and new value mapping is explained in the following.

% % 
% TODO: Explain the format of the raw trace in impl.tex.
% %

@@

\subsection{SSA Form of the Trace}~\label{sec:ssa}
First of all, we rewrite the trace into SSA form, such that every variable in defined exactly once in the trace. This requirement is a prerequisite for the computation of the mapping, in which each variable is mapped to exactly one value. Another side effect is that the def/use chains become explicit in the SSA form, which simplifies the following analysis steps.

\tool\ handles the local assignment and heap accesses differently.
\begin{itemize}
\item {\bf Local Assignment\ } The SSA form for local assignment resembles the SSA form of static instructions in compiler optimization, except that the loops and recursions are fully resolved in a concrete trace, obviating the need for the {\sf Phi} node. More concretely, we replace the variable $v$ defined in an event, as well as the following uses of the definition, to a new variable $v^{id}$, where $id$ is the unique id of the event. The uniqueness of the id guarantees that no two events define the same variable, i.e., each variable is defined exactly once.  Consider the example in Figure~\ref{fig:running2}, where the line number is also used as event id,  the local variable $x$ defined at line 4 is renamed as $x^4$ in the trace, while $x$ defined at line 9 is renamed as $x^9$, which is used in the following two events.  At line 5, where the object $o3$ is created,  we define a reference variable $w^5$.  Internally,  the object $o3$ is encoded as a unique integer that indicates its heap address.
\item {\bf Heap accesses\ } The def/use relation between local heap accesses can be locally determined. Therefore, we treat them in the same way as local assignment. The def/use relations between shared heap accesses are more complex. We may link a read with different writes under different schedules, e.g., the read access to {\tt y} at line 4 in Figure \ref{fig:running} can either obtain the definition before line 1 or at line 1 or at line 3. Therefore, we introduce two symbols to represent the shared read and shared write respectively and use them to establish different links flexibly. The details are as follows. 
\begin{itemize}
\item {\bf Local Heap Accesses\ } Given the write $x.f=y$ to a local heap location at event $e_{id}$,  we introduce a fresh local variable {\tt $l^{id}_{o.f}$} ($o$ is the runtime object referenced by $x$) to replace $x.f$. We also replace the uses of the location $o.f$ accordingly.
Here  the subscript $o.f$ is interpreted as a heap location, i.e., the field $f$ of the object $o$. Consider the example in Figure~\ref{fig:running2}, at lines 6-7, we replace the local heap accesses as the local variable $l^6_{o3.f}$.
%An assumption here is the uses and the definition still refer to the same base object $o$ in the predicted run, which is enforced through the constraint specification %(Section~\ref{sec:relax1}).
\item  {\bf Shared Heap Accesses\ }    Given the write $x.f=y$ or the read $y=x.f$ of a shared location at event $e_{id}$, we introduce two symbols, $W^{id}_{\tt o.f }$ and $R^{id}_{\tt  o.f }$ respectively, where $o$ is the runtime object referenced by the $x$. Similarly, the subscript $o.f$ is interpreted as a heap location, i.e., the field $f$ of the object $o$.
\end{itemize}
We use the heap location such as $o.f$ in the the symbolic form of heap accesses. An underlying assumption is that the heap location remains unchanged in our predictive analysis, which we ensure through additional constraints (Section~\ref{sec:relax1}).
\end{itemize}

@@

\subsection{Constraint System}~\label{sec:constraints}
Based on the SSA form of the original trace $\tau$, we build the constraints to compute a new trace $\tau'$ with the new schedule $O'$ and new value mapping $\theta'$.
Consider the example in Figure~\ref{fig:running2}, in the original trace $\tau$, the event $e_8$ reads the value of 2 from $e_3$, therefore, $e_{11}$ must happen after $e_2$. Existing approaches enforce the same value dependence between $e_3$ and $e_8$, leading to the miss of the race between $e_2$ and $e_{11}$.
Our analysis can derive a new trace $\tau'$ to find the race by relaxing the dependence. In $\tau'$, $e_8$ reads from a different event $e_1$, producing the new value mapping, $\theta'(y^8)=3$. Accordingly, at the event $e_9$, the value mapping for $x^9$ is also updated as 13 ($\theta'(y^8)$=3, $\theta'(z^7)=10$), enabling the true branch at line 10. Finally, the events $e_2$ and $e_{11}$ are scheduled to run concurrently.  The relaxation needs to respect a set of constraints, which we explain in the following. We omit the synchronization constraints intentionally, as they are well explained in all existing predictive analysis techniques~\cite{yannis, pldi14}.

@@

{\bf Race Condition\ } Following the standard definition, pair ($e_i$, $e_j$) of events forms a race iff (1) $e_i$ and $e_j$ are accesses of the same location $\ell$ by different threads, (2) at least one of them writes to $\ell$, and (3) $e_i$ and $e_j$ run concurrently. We refer to the candidate pair throughout this section as (candidate) \emph{racy events}.

We first identify all candidate pairs of events that access the same location from different threads (and at least one is a write). We then check each pair  separately.   The checking is achieved by encoding all necessary constraints and invoking a constraint solver.  The first constraint asserts the feasibility of the concurrent execution of the racy events:
$$
O'(e_i) = O'(e_{j})
$$
In Figure~\ref{fig:running2}, the race condition for the race pair, ($e_2$, $e_{11}$), is $O'(e_2)=O'(e_{11})$

{\bf Intra-thread Constraints\ } Given the candidate race pair, $e_i$ and $e_j$, suppose $e_i$ is after $e_j$ in the original trace $\tau$. The predictive analysis reschedules the events in the prefix of $e_i$, i.e., prior to $e_i$, so that $e_i$ and $e_j$ can run concurrently. We refer to the prefix as $prefix(\tau, e_i)$. Throughout this section, we only consider the events in the prefix.

The predictive analysis by design requires that each thread should follow the same event sequence prior to $e$ as in the original run.  The preservation of the event sequence for each thread requires the following intra-thread constraints:

@@

Without loss of generality, we assume the branch event $e_k$ is in the form of $if(x<y)$. Then  we require that 
$$
\bigwedge_{e_k \in prefix(\tau,e_i) \wedge  inst^{e_k}=if(x<y).\ }  \theta'(x)<\theta'(y) 	\equiv \theta(x)<\theta(y)
$$ 
The constraint specifies that the branch condition should be evaluated as the same boolean value in the predicted run $\tau'$ as in the original trace $\tau$.
 Importantly, Unlike existing analyses \cite{yannis,pldi14}, we do not pose the requirement that the values flowing into branching statements remain the same, but adopt the relaxed requirement that the evaluation of branching expressions remains the same. Consider the example in Figure~\ref{fig:running2}, given the branch if($x^9$>11) at $e_10$, the original mapping is $\theta(x^9)=12$. Existing analyses require the same value mapping for $x^9$  in the predicted run, while we allow  a different value mapping, e.g., $\theta'(x^9)=13$, which retains the same truth value for the branch expression.
\item {\bf Intra-thread Order Constraints\ } The events in the sequence should follow the same order as in the original trace. More formally, we require that
$$
\begin{array}{rl}
\forall e_m, e_n \in prefix(\tau,e_i), s.t., & t^{e_m} = t^{e_n}. \\
 O(e_m) < O(e_n)\  & \Rightarrow O'(e_{m}) < O'(e_{n}) 
\end{array}
$$ 
This constraint specifies that two events from the same thread should follow the same order as reflected by the ids of the events.
\item {\bf Intra-thread Value Constraints\ } The value mapping of the variables should not contradict the value constraints imposed by each instruction.   Without loss of generality, suppose the instruction is a local assignment in the form of $x=y+z$, then we require that
$$
\bigwedge_{e_k \in prefix(\tau,e_i) \wedge inst^{e_k}=x=y+z}\
	\theta'(x)=\theta'(y)+\theta'(z)
$$
Remember that every variable may obtain a different value in the new trace $\tau'$ because it may read from some shared variable affected by the rescheduling.
The constraint specifies that, the variables should be re-assigned to some values that are consistent with the instruction. Consider the example in Figure~\ref{fig:running2}, given the event $x^9=y^8+z^7$ at line 9, if the mappings $\theta'(y^8)$ and $\theta'(z^7)$ change to 3 and 10 respectively, the mapping $\theta'(x^9)$ should change to 13 consistently. In the above, we use the local assignment event for demonstration, in general, the intra-thread value constraints should also be specified for the {\sf  heapr/heapw} events.
\end{itemize}

>>>>

\section{Relaxation of Flow Dependencies}~\label{sec:relax1}
Given the trace, $\tau=<\Gamma , \{\tau_{t_1}, \tau_{t_2}, \dots \tau_{t_n} \}, O, \theta>$, the main goal of our analysis is to derive a new trace,  $\tau'=<\Gamma , \{\tau_{t_1}, \tau_{t_2}, \dots \tau_{t_n} \}, O', \theta'>$, which reschedules the events from different threads (while preserving intra-thread event order) to prove the race. The key insight underlying the first relaxation is that a thread is allowed to read a different value, potentially from a different write, as long as branching decisions remain unchanged.
Note that we also need to compute the new value mapping $\theta'$, because the rescheduling may alter the value every variable is assigned to. The computation of the new schedule and new value mapping is explained in the following.

@@

\subsection{SSA Form of the Trace}~\label{sec:ssa}
First of all, we rewrite the trace into SSA form, such that every variable in defined exactly once in the trace. This requirement is a prerequisite for the computation of the mapping, in which each variable is mapped to exactly one value. Another side effect is that the def/use chains become explicit in  SSA form, which simplifies the following analysis steps.

\tool\ handles the local assignment and heap accesses differently.
\begin{itemize}
\item {\bf Local Assignment\ } The SSA form for local assignment resembles the SSA form of static instructions in compiler optimization, except that the looping and recursion are fully resolved in a concrete trace, obviating the need for {\sf Phi} nodes. More concretely, we replace the variable $v$ defined in an event, as well as the following uses of the definition, to a new variable $v^{id}$, where $id$ is the unique id of the event. The uniqueness of the id guarantees that no two events define the same variable, i.e., each variable is defined exactly once.  In the example in Figure~\ref{fig:running2}, for example,
the local variable $x$ defined at line 4 is renamed as $x^4$.
% where the line number is also used as event id,  the local variable $x$ defined at line 4 is renamed as $x^4$ in the trace, while $x$ defined at line 9 is renamed as $x^9$, which is used in the following two events.  At line 5, where the object $o3$ is created,  we define a reference variable $w^5$.  Internally,  the object $o3$ is encoded as a unique integer that indicates its heap address.
\item {\bf Heap accesses\ } The def/use relation between local heap accesses can be determined locally. Therefore, we treat them in the same way as local assignment. The def/use relations between shared heap accesses are more complex. We may link a read with different writes under different schedules, e.g., the read access to {\tt y} at line 4 in Figure \ref{fig:running} can either obtain the definition before line 1 or at line 1 or at line 3. Therefore, we introduce two symbols to represent the shared read and shared write respectively and use them to establish different links flexibly. The details are as follows. 
\begin{itemize}
\item {\bf Local Heap Accesses\ } Given the write $x.f=y$ to a local heap location at event $e_{id}$,  we introduce a fresh local variable {\tt $l^{id}_{o.f}$} ($o$ is the runtime object referenced by $x$) to replace $x.f$. We also replace the uses of the location $o.f$ accordingly.
Here  the subscript $o.f$ is interpreted as a heap location, i.e., the field $f$ of the object $o$. Consider the example in Figure~\ref{fig:running2}. At lines 6-7, we replace the local heap accesses as the local variable $l^6_{o3.f}$.
%An assumption here is the uses and the definition still refer to the same base object $o$ in the predicted run, which is enforced through the constraint specification %(Section~\ref{sec:relax1}).
\item  {\bf Shared Heap Accesses\ }    Given the write $x.f=y$ or the read $y=x.f$ of a shared location at event $e_{id}$, we introduce two symbols, $W^{id}_{\tt o.f }$ and $R^{id}_{\tt  o.f }$ respectively, where $o$ is the runtime object referenced by $x$. Similarly, the subscript $o.f$ is interpreted as a heap location, i.e., the field $f$ of the object $o$.
\end{itemize}
We use the heap location such as $o.f$ in the the symbolic form of heap accesses. An underlying assumption is that the heap location remains unchanged in our predictive analysis, which we ensure through additional constraints (Section~\ref{sec:relax1}).
\end{itemize}

@@

\subsection{Constraint System}~\label{sec:constraints}
Based on the SSA form of the original trace $\tau$, we build the constraints to compute a new trace $\tau'$ with the new schedule $O'$ and new value mapping $\theta'$.
Consider the example in Figure~\ref{fig:running2}. In the original trace $\tau$, the event $e_8$ reads the value of 2 from $e_3$. Therefore, $e_{11}$ must happen after $e_2$. Existing approaches enforce the same value dependence between $e_3$ and $e_8$, thereby missing the race between $e_2$ and $e_{11}$.
Our analysis allows event $e_8$ to read from $e_1$, producing the new value mapping $\theta'(y^8)=3$. Accordingly, at $e_9$ the value mapping for $x^9$ is also updated as 13 ($\theta'(y^8)$=3, $\theta'(z^7)=10$), enabling the true branch at line 10. Finally, the events $e_2$ and $e_{11}$ are scheduled to run concurrently.  The relaxation needs to respect a set of constraints, which we explain in the following. We omit the synchronization constraints intentionally, as they are well explained in all existing predictive analysis techniques~\cite{yannis, pldi14}.

@@

{\bf Race Condition\ } Following the standard definition, pair ($e_i$, $e_j$) of events forms a race iff (1) $e_i$ and $e_j$ access a mutual  location $\ell$, (2) at least one of them writes $\ell$, and (3) $e_i$ and $e_j$ run concurrently (being associated with different threads). We refer to the candidate pair throughout this section as (candidate) \emph{racy events}.

We first identify all candidate pairs of events that access the same location from different threads (and at least one is a write). We then check each pair  separately.   The checking is achieved by encoding all necessary constraints.  The first constraint asserts the feasibility of the concurrent execution of the racy events: $O'(e_i) = O'(e_{j})$.
%In Figure~\ref{fig:running2}, the race condition for the race pair, ($e_2$, $e_{11}$), is $O'(e_2)=O'(e_{11})$

{\bf Intra-thread Constraints\ } Given racy events $e_i$ and $e_j$, suppose $e_i$ is after $e_j$ in the original trace $\tau$. The predictive analysis reschedules the events in the prefix up to $e_i$, so that $e_i$ and $e_j$ can run concurrently. We refer to the prefix as $prefix(\tau, e_i)$. Throughout this section, we only consider the events in the prefix.

Preservation of the event sequence (up to $e$) for each thread requires the following intra-thread constraints:

@@

Without loss of generality, we assume the branch event $e_k$ is in the form of $if(x<y)$. Then  we require that 
$$
\bigwedge_{e_k \in prefix(\tau,e_i) \wedge  inst^{e_k}=if(x<y).\ }  \theta'(x)<\theta'(y) 	\equiv \theta(x)<\theta(y)
$$ 
The constraint specifies that the branch condition should be evaluated as the same boolean value in the predicted run $\tau'$ as in the original trace $\tau$.
 Importantly, Unlike existing analyses \cite{yannis,pldi14}, we do not pose the requirement that the values flowing into branching statements remain the same, but adopt the relaxed requirement that the evaluation of branching expressions remains the same. Consider the example in Figure~\ref{fig:running2}, given the branch if($x^9$>11) at $e_10$, the original mapping is $\theta(x^9)=12$. We allow  a different value mapping, e.g., $\theta'(x^9)=13$, which retains the same truth value for the branch expression.
\item {\bf Intra-thread Order Constraints\ } The events in the sequence should follow the same order as in the original trace. More formally, we require that
$$
\begin{array}{rl}
\forall e_m, e_n \in prefix(\tau,e_i), s.t., & t^{e_m} = t^{e_n}. \\
 O(e_m) < O(e_n)\  & \Rightarrow O'(e_{m}) < O'(e_{n}) 
\end{array}
$$ 
This constraint specifies that two events from the same thread should follow the same order as reflected by the ids of the events.
\item {\bf Intra-thread Value Constraints\ } The value mapping of the variables should not contradict the value constraints imposed by each instruction.   Without loss of generality, suppose the instruction is a local assignment in the form of $x=y+z$, then we require that
$$
\bigwedge_{e_k \in prefix(\tau,e_i) \wedge inst^{e_k}=x=y+z}\
	\theta'(x)=\theta'(y)+\theta'(z)
$$
Remember that every variable may obtain a different value in the new trace $\tau'$ because it may read from some shared variable affected by the rescheduling.
The constraint specifies that, the variables should be re-assigned to some values that are consistent with the instruction. Consider the example in Figure~\ref{fig:running2}, given the event $x^9=y^8+z^7$ at line 9, if the mappings $\theta'(y^8)$ and $\theta'(z^7)$ change to 3 and 10 respectively, the mapping $\theta'(x^9)$ should change to 13 consistently. In the above, we use the local assignment event for demonstration, in general, the intra-thread value constraints should also be specified for the {\sf  heapr/heapw} events.
\end{itemize}
